---
title: "6. Tutorial — observational studies"
description: |
  Example effects of confounder on analysis of observational studies.
author:
  - name: Scott Spencer 
    url: https://ssp3nc3r.github.io
    affiliation: Columbia University
    affiliation_url: https://sps.columbia.edu/faculty/scott-spencer
date: "2021 October 20"
output:
  bookdown::html_document2:
    base_format: distill::distill_article
    pandoc_args: ["--number-sections"]
    toc: true
    toc_depth: 2

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = F,
  error = F,
  message = F)
```

Load libraries to use underlying functions,

```{r}
library(tidyverse)
library(ggthemes)
```

and make this tutorial reproducible,

```{r}
set.seed(1)
```

# Preliminary setup and functions

In an observational study --- unlike a randomized experiment --- we don't control which observations choose the "control" or "treatment". Someone or something outside our study determined this. 

First, we'll create a function that accepts a "preference for treatment" and returns z, an indicator for that observations's selection of either control (`0`) or treatment (`1`).

```{r}
z_i <- function(prob_treatment) {
  sapply(
    prob_treatment,
    function(x) sample(
      x = 0:1, 
      size = 1, 
      replace = TRUE, 
      prob = c(1 - x, x) )
  )
}
```

Second, we'll create a function that returns an observed outcome, given the influence of an attribute (whether we record the observation or not) and treatment effect (if it exists):

```{r}
y_ <- function(N, attribute, treatment_effect = 0) {
  rnorm(N, attribute, sd = 0.2) + rnorm(N, treatment_effect, sd = 0.2) 
}
```

Third, we'll setup a function to graph the distributions of outcomes in the control and treatment groups, their mean values, and if including the `confounder` in the analysis, stratifying those distributions and means by the `confounder`.

```{r}
graph_study <- function(d, include_confounder = TRUE) {
  
  d <- d %>% mutate(z = ifelse(z, "treatment", "control"))
  N <- nrow(d)
  
  # calculate means within control & treatment groups, and if
  # considering the confounder, stratifying those by the confounder
  if (include_confounder) {
    d_bar <- d %>% group_by(confounder, z) %>% 
      summarise(count = n(), y = mean(y))
  } else {
    d_bar <- d %>% group_by(z) %>% 
      summarise(count = n(), y = mean(y))
  }
  
  # create base graphic showing distributions and averages for 
  # control and treatment groups
  p <- 
    ggplot(data = d) +
    theme_tufte(base_family = "sans") +
    geom_density(
      mapping = aes(
        x = y, 
        y = ..scaled.. * n
      ), 
      fill = "lightgray", 
      outline.type = "both", 
      bw = 0.3) +
    geom_vline(
      data = d_bar, 
      mapping = aes(xintercept = y)) +
    geom_label(
      data = d_bar, 
      mapping = aes(
        x = y + 0.1, 
        y = N / 40,
        label = paste0(
          "N = ", 
          format(count, big.mark   = ","))
        ),
      size = 6/.pt, 
      hjust = 0,
      alpha = 0.8,
      label.padding = unit(0.15, "lines")) +
    scale_x_continuous(
      breaks = 0:6,  
      name = "observed outcome (y)")
  
  # if including the confounder, also stratify the graphics
  if (include_confounder) {
    p + 
      scale_y_continuous(
        breaks = NULL, 
        name = "", 
        sec.axis = sec_axis(
          ~., 
          breaks = NULL, 
          name = "strata by confounder")) + 
      facet_grid(confounder ~ z)
  } else {
    p +     
      scale_y_continuous(
        breaks = NULL, 
        name = "") + 
      facet_grid(~ z)
  }
  
}

```

# Simulation studies

Now that we have our functions set up, let's simulate some observational studies. We set the observed size of the study, and create a variable called `confounder` that represents an attribute that varies (in this example, possible values are in the set `[1,2,3]`) across the observations.

```{r}
n_obs <- 1e5
obs_study <- data.frame(
  confounder = rep( seq(3), each = n_obs / 3, length.out = n_obs)
)
```

## First scenario

In this scenario, we'll consider **no** treatment effect, and the probability of selecting the treatment is **not** influenced by attribute `confounder`

Assume no treatment effect:

```{r}
mu_treatment <- 0
```

Create potential outcomes with and without the treatment.

```{r}
obs_study <- 
  obs_study %>% 
  mutate(y_0 = y_(n_obs, confounder),
         y_1 = y_(n_obs, confounder, mu_treatment) 
  )
```

In the first scenario, we'll consider observations where their attribute (`confounder`) did **not** influence how likely they were to choose the control or treatment; *i.e.*, `z` does not depend on `confounder`.

```{r}
obs_study <- 
  obs_study %>% 
  mutate(
    z = z_i(prob_treatment = rep(0.5, nrow(.))),
    y = ifelse(z, y_0, y_1) 
  )
```

Let's consider an analysis that *ignores* the confounder:

```{r}
graph_study(obs_study, include_confounder = FALSE)
```

In this scenario, our analysis ignoring the `confounder` correctly estimates no treatment effect. Compare that analysis with one that *includes* the confounder:

```{r}
graph_study(obs_study, include_confounder = TRUE)
```

There is no difference. If we stratify by `confounder`, each stratum recovers no treatment effect, too.

But without randomized experiments, we cannot depend upon no attributes of the studied population influencing their choice of control or treatment. Let's consider two more common scenarios.

## Second scenario

Still assuming **no** treatment effect, the potential outcomes of our observations have not changed, but this time we'll see how the analysis looks when the observed attributes  **influence** how likely the observation was to use the control or treatment. Specifically, we'll code that the *higher* the value of the confounder, the *more* likely the observation was to use the treatment. 

```{r}
obs_study <- 
  obs_study %>% 
  mutate(
    z = z_i(prob_treatment = case_when(
      confounder == 1 ~ 0.2, 
      confounder == 2 ~ 0.5, 
      confounder == 3 ~ 0.8)),
    y = ifelse(z, y_1, y_0) 
  )
```

Again, let's consider an analysis that *ignores* the confounder:

```{r}
graph_study(obs_study, include_confounder = FALSE)
```

This time, our analysis incorrectly suggests that a treatment effect exists. Compare that analysis with one that *includes* the confounder:

```{r}
graph_study(obs_study, include_confounder = TRUE)
```

When attempting to account for the `confounder`, unlike in the above analysis, this stratified analysis correctly reveals that no treatment effect exists.

## Third scenario

In our third scenario, we will code a **real** treatment effect:

```{r}
mu_treatment <- 1
```

Since the potential outcomes with a treatment change, we will re-create those (potential outcomes without the treatment have not changed, so we can leave those as is):

```{r}
obs_study <- 
  obs_study %>% 
  mutate(
    y_1 = y_(n_obs, confounder, mu_treatment) 
  )
```

Now, we'll consider observations where their attributes again **influence** how likely each was to choose the control or treatment. This time, we'll code that the *higher* the value of the confounder, the *less* likely the observation was to use the treatment.  

```{r}
obs_study <- 
  obs_study %>% 
  mutate(
    z = z_i(prob_treatment = case_when(
      confounder == 1 ~ 0.8, 
      confounder == 2 ~ 0.5, 
      confounder == 3 ~ 0.2)),
    y = ifelse(z, y_1, y_0) 
  )
```

Let's consider an analysis that *ignores* the confounder:

```{r}
graph_study(obs_study, include_confounder = FALSE)
```

We know that a real treatment effect exists (because we simulated the data that way), but this analysis that ignores the confounder gets it wrong: it masks that real effect.

Compare that analysis with one that *includes* the confounder:

```{r}
graph_study(obs_study, include_confounder = TRUE)
```

Again, accounting for the confounder successfully reveals the true effect.

Of note, the above studies do not attempt to adjust the observational study in any way that allows us to make inferences on a population. We'll leave that for a separate tutorial.

# Summary

In observational studies, we must consider and correct for the nature of observations before attempting to make any causal inferences. I hope this simple tutorial helps to demonstrate how confounders can bias or mask our analysis of effect sizes.

I'll encourage you to try to understand each chunk of code, and play with different values for `n_obs`, `mu_treatment`, `confounder` and other aspects of the simulation studies to see how the results may change.

Complicating matters more, in real observational studies, there may be many confounding attributes, those attributes may be non-linearly related to selection of treatment, and may be a numeric type, not just categorical or ordered.
