---
title: "5. Research Design"
description: |
  Example code used in class discussion.
author:
  - name: Scott Spencer 
    url: https://ssp3nc3r.github.io
    affiliation: Columbia University
    affiliation_url: https://sps.columbia.edu/faculty/scott-spencer
date: "2021 October 13"
output:
  bookdown::html_document2:
    base_format: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = F,
  error = F,
  message = F)
```

Load libraries to use underlying functions

```{r}
library(tidyverse)
library(ggthemes)
library(latex2exp)
library(patchwork)
theme_set( theme_tufte(base_family = "sans") )
```

Here's our student-t functions from last time:

```{r}
dstudent_t <- 
  function(x, df, mu = 0, sigma = 1, log = FALSE) {
  if (log) {
    dt( (x - mu) / sigma, df = df, log = TRUE ) - log(sigma)
  } else {
    dt( (x - mu) / sigma, df = df ) / sigma
  }
}

pstudent_t <- 
  function(q, df, mu = 0, sigma = 1, lower.tail = TRUE, log.p = FALSE) {
  pt( (q - mu) / sigma, df = df, lower.tail = lower.tail, log.p = log.p )
}

qstudent_t <- 
  function(p, df, mu = 0, sigma = 1, lower.tail = TRUE, log.p = FALSE) {
  qt( p, df = df, lower.tail = lower.tail, log.p = log.p ) * sigma + mu
}

rstudent_t <- 
  function(n, df, mu = 0, sigma = 1) {
  rt( n, df = df ) * sigma + mu
}
```

# Slide 10

Simulate data so we can see how the test works.

```{r}
set.seed(1)
# population
mu1 <- 4
pop1 <- rnorm(1e5, mean = mu1, sd = 2)

# sample from the populations
n1 <- 50
x1 <- sample(x = pop1, size = n1, replace = FALSE)
```

Run the test.

```{r}
# setup the test to calculate manually
xbar1 <- mean(x1)
s1 <- sd(x1)
nu <- n1 - 1

# pretend we know population mu but not sigma
t <- (xbar1 - mu1) / (s1 / sqrt(n1))

# manually get p-value
p <- pstudent_t(q = t, df = nu)
p
```

Consider the results graphically.

```{r}
cdf <- 
  ggplot() +
  theme(axis.line.x = element_line()) +
  scale_x_continuous(breaks = seq(-3, 3)) +
  stat_function(
    fun = pstudent_t,
    args = list(df = nu, mu = 0, sigma = 1),
    stat = "density",
    xlim = c(-3, 3)
  ) +
  geom_vline(xintercept = t, color = "dodgerblue") +
  annotate("text", x = t + 0.05, y = 0, label = TeX("$t$") , color = "dodgerblue", size = 10/.pt, hjust = 0, vjust = 0) +
  
  geom_hline(yintercept = p, color = "pink") +
  annotate("text", x = -4, y = p + 0.01, label = TeX("$p$") , color = "pink", size = 10/.pt, hjust = 0, vjust = 0) +
  
  geom_hline(yintercept = 0.05, color = "gray", linetype = "dotted") +
  annotate("text", x = -4, y = 0.06, label = TeX("$\\alpha$, arbitrary threshold") , color = "gray", size = 10/.pt, hjust = 0, vjust = 0) +
  labs(x = TeX("$\\sigma$"), y = TeX("$F_t(x, \\nu)$"))

cdf
```

Compare our calculations against r base function.

```{r}
a <- t.test(x = x1, mu = mu1, alternative = "less", conf.level = 0.95)
p == a$p.value
```

# Slide 12

Simulate second population for next example.

```{r}
# second population
mu2 <- 2
pop2 <- rnorm(1e5, mean = mu2, sd = 3)

# sample from the second population
n2 <- 50
x2 <- sample(x = pop2, size = n2, replace = FALSE)
```

Run the test.

```{r}
xbar2 <- mean(x2)
s2 <- sd(x2)
  
t <- ( xbar1 - xbar2 ) / sqrt( s1 ^ 2 / n1 + s2 ^ 2 / n2)
  
nu <- ( s1 ^ 2 / n1 + s2 ^ 2 / n2 ) ^ 2 / 
      ( ( s1 ^ 2 / n1 ) ^ 2 / (n1 - 1) + ( s2 ^ 2 / n2 ) ^ 2 / (n2 - 1) )

p <- pstudent_t(q = t, df = nu)  
p
```

Graphically consider the results.

```{r}
cdf <- 
  ggplot() +
  theme(axis.line.x = element_line()) +
  scale_x_continuous(breaks = seq(-3, 3)) +
  stat_function(
    fun = pstudent_t,
    args = list(df = nu, mu = 0, sigma = 1),
    stat = "density",
    xlim = c(-3, 3)
  ) +
  geom_vline(xintercept = t, color = "dodgerblue") +
  annotate("text", x = t + 0.05, y = 0, label = TeX("$t$") , color = "dodgerblue", size = 10/.pt, hjust = 0, vjust = 0) +
  
  geom_hline(yintercept = p, color = "pink") +
  annotate("text", x = -4, y = p + 0.01, label = TeX("$p$") , color = "pink", size = 10/.pt, hjust = 0, vjust = 0) +
  
  geom_hline(yintercept = 0.05, color = "gray", linetype = "dotted") +
  annotate("text", x = -4, y = 0.06, label = TeX("$\\alpha$, arbitrary threshold") , color = "gray", size = 10/.pt, hjust = 0, vjust = 0) +
  labs(x = TeX("$\\sigma$"), y = TeX("$F_t(x, \\nu)$"))

cdf
```

Compare our calculations against the base R function.

```{r}
a <- t.test(x1, x2, alternative = "less", var.equal = FALSE, conf.level = 0.95)
p == a$p.value
```

# Slide 14

Create example with proportions.

```{r}
# population proportion
pi <- 0.4

# population of proportions
pop1 <- rbinom(n = 1e5, size = 1, prob = pi)

# observed proportion (sample or experiment)
p1 <- sample(pop1, size = n1)
phat1 <- mean(p1)
```

Calculate the test.

```{r}
# calculate test statistic
z <- ( phat1 - pi ) / sqrt( pi * (1 - pi) / n1 )

# get location on cdf of standard normal distribution
p <- pnorm(q = z)
```

Visualize the results.

```{r}
cdf <- 
  ggplot() +
  theme(axis.line.x = element_line()) +
  scale_x_continuous(breaks = seq(-3, 3)) +
  stat_function(
    fun = pnorm,
    args = list(mean = 0, sd = 1),
    stat = "density",
    xlim = c(-3, 3)
  ) +
  geom_vline(xintercept = z, color = "dodgerblue") +
  annotate("text", x = z + 0.05, y = 0, label = TeX("$z$") , color = "dodgerblue", size = 10/.pt, hjust = 0, vjust = 0) +
  
  geom_hline(yintercept = p, color = "pink") +
  annotate("text", x = -4, y = p + 0.01, label = TeX("$p$") , color = "pink", size = 10/.pt, hjust = 0, vjust = 0) +
  
  geom_hline(yintercept = 0.05, color = "gray", linetype = "dotted") +
  annotate("text", x = -4, y = 0.06, label = TeX("$\\alpha$, arbitrary threshold") , color = "gray", size = 10/.pt, hjust = 0, vjust = 0) +
  labs(x = TeX("$\\sigma$"), y = TeX("$F_\\phi(z)$"))

cdf
```

# Slide 17

Create two samples from different distributions.

```{r}
# simulate samples from experiment, samples from different distributions
n <- 10

set.seed(1)

sample1 <- rbeta(n, 2, 2)
sample2 <- rbeta(n, 2, 5)
```

Calculate the test.

```{r}
# organize data
d <- 
  data.frame(
    sample = rep(1:2, each = n),
    values = c(sample1, sample2)
  ) %>%
  arrange(values) %>%
  mutate(rank = seq( n() )) %>%
  group_by(values) %>%
  mutate(rank = mean(rank))


Ta <- filter(d, sample == 1) %>% .$rank %>% sum()
Tb <- filter(d, sample == 2) %>% .$rank %>% sum()  

n1 <- with(d, sum(sample == 1))
n2 <- with(d, sum(sample == 2))

Ua <- n1 * n2 + (n1 * (n1 + 1)) / 2 - Ta 
Ub <- n1 * n2 + (n2 * (n2 + 1)) / 2 - Tb
U <- min(Ua, Ub)

EU <-  n1 * n2 / 2
sigma <- sqrt( n1 * n2 * (n1 + n2 + 1) / 12 )

z <- (U - EU) / sigma
p <- pnorm(z)
p
```


# Slide 18

Compare our results to a base R function.

```{r}
a <- wilcox.test(x = sample1, y = sample2, 
            correct = FALSE, exact = FALSE,
            alternative = "greater")
p == a$p.value
```

# Slide 19

Visualize both the population distribuitons,

```{r}
dist1 <- ggplot() +
  theme(axis.line.x = element_line()) +
  scale_x_continuous(breaks = seq(0, 1)) +
  scale_y_continuous(breaks = NULL) +
  stat_function(
    fun = dbeta,
    args = list(shape1 = 2, shape2 = 2),
    stat = "density",
    xlim = c(0, 1)
  ) +
  labs(y = "density", title = "Beta(2, 2)")

dist2 <- ggplot() +
  theme(axis.line.x = element_line()) +
  scale_x_continuous(breaks = seq(0, 1)) +
  scale_y_continuous(breaks = NULL) +
  stat_function(
    fun = dbeta,
    args = list(shape1 = 2, shape2 = 5),
    stat = "density",
    xlim = c(0, 1)
  ) +
  labs(y = "", title = "Beta(2, 5)")

dist1 + dist2
```

and our results,

```{r}
cdf <- 
  ggplot() +
  theme(axis.line.x = element_line()) +
  scale_x_continuous(breaks = seq(-3, 3)) +
  stat_function(
    fun = pnorm,
    args = list(mean = 0, sd = 1),
    stat = "density",
    xlim = c(-3, 3)
  ) +
  geom_vline(xintercept = z, color = "dodgerblue") +
  annotate("text", x = z + 0.05, y = 0, label = TeX("$z$") , color = "dodgerblue", size = 10/.pt, hjust = 0, vjust = 0) +
  
  geom_hline(yintercept = p, color = "pink") +
  annotate("text", x = -4, y = p + 0.01, label = TeX("$p$") , color = "pink", size = 10/.pt, hjust = 0, vjust = 0) +
  
  geom_hline(yintercept = 0.05, color = "gray", linetype = "dotted") +
  annotate("text", x = -4, y = 0.06, label = TeX("$\\alpha$, arbitrary threshold") , color = "gray", size = 10/.pt, hjust = 0, vjust = 0) +
  labs(x = TeX("$\\sigma$"), y = TeX("$F_\\phi(z)$"))

cdf
```

# Slide 21

Example — test equality of proportions of male and female applicants to Berkeley. Let's setup the data and calculate our test statistic and p-value.

```{r}
data(UCBAdmissions)

d_stat <- 
  
  UCBAdmissions %>% 
  
  as.data.frame() %>%
  
  group_by(Gender) %>%
  
  summarise(O = sum(Freq)) %>%
  
  ungroup() %>%
  
  mutate(E = mean(O)) %>%
  
  summarise(
    w = sum((O - E) ^2 / E),
    nu = n_distinct(Gender) - 1
  ) %>%
  
  mutate(
    p = pchisq(w, nu, lower.tail = FALSE)
  )
```

Let's graph our result.

```{r}
cdf <- 
  ggplot() +
  theme(axis.line.x = element_line()) +
  scale_x_continuous(breaks = seq(0, 200, by = 50)) +
  stat_function(
    fun = pchisq,
    args = list(df = d_stat$nu),
    stat = "density",
    xlim = c(0, 200)
  ) +
  geom_vline(xintercept = d_stat$w, color = "dodgerblue") +
  annotate("text", x = d_stat$w + 1, y = 0, label = TeX("$w$") , color = "dodgerblue", size = 10/.pt, hjust = 0, vjust = 0) +
  
  geom_hline(yintercept = d_stat$p, color = "pink") +
  annotate("text", x = -4, y = d_stat$p + 0.01, label = TeX("$p$") , color = "pink", size = 10/.pt, hjust = 0, vjust = 0) +
  
  geom_hline(yintercept = 0.05, color = "gray", linetype = "dotted") +
  annotate("text", x = -4, y = 0.06, label = TeX("$\\alpha$, arbitrary threshold") , color = "gray", size = 10/.pt, hjust = 0, vjust = 0) +
  labs(x = "", y = TeX("$F_\\chi^2(z)$"))

cdf
```

# Slide 22

Example — H0 : P(Admit | Gender) = P(Admit) and P(Gender | Admit) = P(Gender)

```{r}
d_stat <- 
  
  UCBAdmissions %>% 
  
  as.data.frame() %>%
  
  mutate(
    Admit_pct = sum(ifelse(Admit == "Admitted", Freq, 0) ) / sum(Freq)
  ) %>%
  
  group_by(Gender) %>%
  
  mutate(
    E = sum(Freq) * ifelse(Admit == "Admitted", Admit_pct, 1 - Admit_pct)
  ) %>%
  
  group_by(Gender, Admit) %>%
  
  summarise(
    O = sum(Freq), 
    E = mean(E)
  ) %>%
  
  ungroup() %>%
  
  summarise(
    w = sum((O - E)^2 / E),
    nu = (n_distinct(Admit) - 1) * (n_distinct(Gender) - 1)
  ) %>%
  
  mutate(
    p = pchisq(w, nu, lower.tail = FALSE)
  )
```

Let's graph our result.

```{r}
cdf <- 
  ggplot() +
  theme(axis.line.x = element_line()) +
  scale_x_continuous(breaks = seq(0, 100, by = 50)) +
  stat_function(
    fun = pchisq,
    args = list(df = d_stat$nu),
    stat = "density",
    xlim = c(0, 100)
  ) +
  geom_vline(xintercept = d_stat$w, color = "dodgerblue") +
  annotate("text", x = d_stat$w + 1, y = 0, label = TeX("$w$") , color = "dodgerblue", size = 10/.pt, hjust = 0, vjust = 0) +
  
  geom_hline(yintercept = d_stat$p, color = "pink") +
  annotate("text", x = -4, y = d_stat$p + 0.01, label = TeX("$p$") , color = "pink", size = 10/.pt, hjust = 0, vjust = 0) +
  
  geom_hline(yintercept = 0.05, color = "gray", linetype = "dotted") +
  annotate("text", x = -4, y = 0.06, label = TeX("$\\alpha$, arbitrary threshold") , color = "gray", size = 10/.pt, hjust = 0, vjust = 0) +
  labs(x = "", y = TeX("$F_\\chi^2(z)$"))

cdf
```

